Generalized Bradley-Terry models

### 2025-03-19 inst/jss_hyper3_real.Rnw
cite Plackett

\begin{equation}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
  \end{equation}

@@

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt soccerworld}}
\end{table}

@@

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce likelihood function for this observation:

>>>>

\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
  \end{equation}

@@

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}

@@

Equation~\ref{mostgeneral} is very general and may be applied to a
range of situations in which standard Bradley-Terry strengths are
modified in some way.  In the following section, I apply it to
Plackett-Luce likelihoods~\citep{plackett1975} likelihood function for
rank data.

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975} likelihood function for
this observation:
